---
title: "Data Visualizations Short Paper"
author: "Ratul Esrar"
date: "10/22/2017"
output:
  html_document: default
  pdf_document: default
---

```{r echo=FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(reshape2)
library(zoo)
library(ggrepel)
library(readxl)
require(forcats)
library(rgdal)
library(maptools)
if (!require(gpclib)) install.packages("gpclib", type="source")
gpclibPermit()


# Set working directory
setwd('/Users/ratulesrar/Desktop/data_viz_pbpl/hw2/')

# Import pricestats data from MIT BPP
pricestats <- read_csv('pricestats_bpp_arg_usa.csv')
pricestats_usa <- pricestats %>% 
  filter(country == 'USA') %>%
  mutate(date = as.Date(date, '%d%b%Y')) %>%
  mutate(qtr = as.yearqtr(date)) %>%
  group_by(qtr) %>%
  summarize(avg_quarterly_cpi = mean(indexPS))

# Import OECD CPI data
oecd_cpi <- read_csv('us_cpi_oecd.csv')
oecd_cpi <- oecd_cpi %>%
  filter(Subject == 'Consumer prices - all items', MEASURE == 'IXOB') %>%
  mutate(qtr = as.yearqtr(Time, format = 'Q%q-%Y'))

# Merge the pricestats and oecd data
binded_cpi <- bind_rows(pricestats_usa, oecd_cpi)
agg_cpi <- binded_cpi %>% mutate(cpi = ifelse(is.na(avg_quarterly_cpi), Value, avg_quarterly_cpi)) %>%
  mutate(source = ifelse(is.na(Measure), 'Pricestats', 'OECD')) %>%
  mutate(qtr = as.yearqtr(qtr)) %>%
  select(qtr, cpi, source)

merged_cpi <- left_join(pricestats_usa, oecd_cpi, by='qtr')
merged_cpi <- merged_cpi %>%
  mutate(cpi_oecd=Value, cpi_ps=avg_quarterly_cpi, qtr=as.yearqtr(qtr)) %>%
  mutate(difference=cpi_ps-cpi_oecd) %>%
  select(qtr, cpi_oecd, cpi_ps, difference)

# Import Quarterly Income Data from Bureau of Economic Analysis
income <- read_csv('personal_income_quarterly_bystate_2007-2017.csv', skip = 4)
quarter_cols <- c('2008:Q3', '2015:Q3')
drop_rows <- c('New England', 'Mideast', 'Great Lakes', 'Plains', 'Southeast', 
               'Southwest', 'Rocky Mountain', 'Far West', 'United States')
state_personal_income <- income %>%
  filter(LineCode==1, !GeoName %in% drop_rows) %>%
  select(GeoName, quarter_cols) 
state_personal_income <- state_personal_income %>%
  mutate(state=as.character(GeoName),
         t1=as.integer(state_personal_income[['2008:Q3']]), 
         t2=as.integer(state_personal_income[['2015:Q3']]),
         pct_change=100*(t2-t1)/t1) %>%
  select(state, t1, t2, pct_change) %>%
  mutate(state=factor(state, levels=unique(state)))

quarter_cols_2 <- c('2010:Q1', '2010:Q2', '2010:Q3', '2010:Q4',
                  '2011:Q1', '2011:Q2', '2011:Q3', '2011:Q4',
                  '2012:Q1', '2012:Q2', '2012:Q3', '2012:Q4',
                  '2013:Q1', '2013:Q2', '2013:Q3', '2013:Q4',
                  '2014:Q1', '2014:Q2', '2014:Q3', '2014:Q4',
                  '2015:Q1', '2015:Q2', '2015:Q3')
drop_rows_2 <- c('New England', 'Mideast', 'Great Lakes', 'Plains', 'Southeast', 
                 'Southwest', 'Rocky Mountain', 'Far West')
capita_income <- income %>%
  filter(LineCode==3, !GeoName %in% drop_rows_2) %>%
  select(GeoName, quarter_cols_2) 
state_capita_income <- capita_income %>%
  mutate(state=as.character(GeoName)) %>%
  mutate(state=factor(state, levels=unique(state))) %>%
  select(state, quarter_cols_2)

# Import income distribution data from Urban Institute
income_dist <- read_xlsx('income_dist.xlsx')

# Create theme
theme_dv <- function () {
  theme_bw(base_size=11, base_family='sans') %+replace%
    theme(
      panel.background=element_blank(),
      plot.background=element_rect(fill='gray96', color=NA), 
      legend.background=element_rect(fill='transparent', color=NA),
      legend.key=element_rect(fill='transparent', color=NA))
}

dv_theme_pal <- function() {
    function(n) {
        colors <-c('blue', 'red', 'green')
        unname(colors[seq_len(n)])
    }
}

scale_color_dv <- function(...) {
    discrete_scale('color', 'economist', dv_theme_pal(), ...)
}

scale_fill_dv <- function(...) {
    discrete_scale('fill', 'economist', dv_theme_pal(), ...)
}

```

### Graph 1: Are federally reported inflation rates in the US underestimated?
```{r echo=FALSE, warning=FALSE, message=FALSE, fig.align='center'}
# Plot Graph 1
agg_cpi %>% 
  mutate(label=if_else(qtr==max(qtr), as.character(source), NA_character_)) %>%
  ggplot(aes(x=qtr, y=cpi, color=source)) +
  geom_line() +
  scale_colour_discrete(guide=FALSE) +
  labs(title='Are federally reported inflation rates in the US underestimated?',
       subtitle='Comparing Pricestats and OECD CPI from 2008Q3 - 2015Q3',
       caption='Source: MIT Billion Prices Project, OECD & BLS', 
       x='Year', y='Consumer Price Index (CPI)') +
  geom_label_repel(aes(label=label), na.rm=TRUE) +
  theme_dv()
```

Using daily consumer price index calculations calculated by the Pricestats database is arguably a more precise measure of inflation than quarterly reports by the Bureau of Labor Statistics. Looking at quarterly CPI reports relative to the average of daily CPI calculations per quarter from Pricestats shows an interesting relationship, especially during times of recession. The Pricestats CPI is generally higher than the OECD report except during recession periods.

### Graph 2: The BLS underestimates CPI during recessions

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align='center'}
# Plot Graph 2
merged_cpi %>%
  ggplot(aes(x=qtr, y=difference)) +
  geom_point(alpha=0.65, color='#800000') +
  geom_smooth(se=FALSE, show.legend=TRUE, span=1.5, 
              size=0.5, alpha=0.5, aes(color='Trendline (Pricestats - BLS)')) +
  geom_hline(yintercept=0, aes(linetype='Zero Line'), linetype='solid') +
  labs(title='The BLS underestimates CPI during recessions',
       subtitle='Comparing the difference between Pricestats and OECD CPI from 2008Q3 - 2015Q3',
       caption='Source: MIT Billion Prices Project, OECD & BLS', 
       x='Year', y='Difference in CPI') +
  scale_colour_manual(name='Legend', values='#14bde8') +
  annotate('rect', fill='gray', alpha=0.35, 
           xmin=2008, xmax=2009+6/12, ymin=-Inf, ymax=Inf) +
  theme_dv() +
  theme(legend.position = c(0.83, 0.1))
```

Looking more closely at the differences between the OECD's reported CPI, which is taken from the BLS, and the Pricestats CPI database, we see two interesting things. First, the CPI is trending upward and that makes sense as the United States has not been in a period of deflation since 2008Q3. Second, the difference between Pricestats CPI and OECD CPI is negative during the 2008 recession but has been positive in almost every period since then. It is possible that the BLS has incentive to overestimate CPI during periods of economic stagnation to artificially indicate economic growth.

### Graph 3: Changes in Personal Income Outpaced Change in Inflation

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align='center'}
state_personal_income %>%
  ggplot(aes(x=reorder(x=state, -pct_change), y=pct_change)) + 
  geom_point(color='#800000', alpha=0.65, stat='identity', size=0.75) +
  geom_hline(yintercept=9.8, aes(linetype='Percent Change in CPI'),
             size=0.5, alpha=0.85, linetype='dashed') +
  labs(title='Changes in Personal Income Outpaced Change in Inflation',
     subtitle='Percent Change calculated using 2015Q3 and 2008Q3 data',
     caption='Source: Bureau of Economic Analysis (US Dept of Commerce)', 
     x='State', y='Percent Change in Personal Income') + 
  annotate('text', 'Nevada', 12.5, label='Percent Change in CPI from 2008Q3 to 2015Q3') +
  scale_y_continuous(labels = function(x){paste0(x, '%')}) +
  theme_dv() +
  theme(axis.text.x=element_text(angle=90, hjust=1),
        legend.position=c(0,0))
```

When examining difference in CPI over time, it is also prudent to look at changes to income, or in this case, aggregate personal income by state (seasonally adjusted). Over the time period of this data, the median change in total personal income by state was around 23 percent, while the CPI only increased by 10 percent in the same period of time. It's possible that total personal income is skewed due to the top earners in each state, which may be why that change seems much larger than the change in inflation. 

### Graph 4: Income Distribution Has Not Kept Pace with CPI

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align='center'}
income_dist %>%
  ggplot() +
  geom_line(aes(x=year, y=avg_income, color=percentile)) +
  facet_wrap(~percentile) +
  guides(color=FALSE) +
  scale_y_continuous(labels = function(x){paste0(x, '$')}) +
  labs(title='Family Income Has Not Kept Pace with CPI',
       subtitle='Buying power for families in the 10th and 50th Percentiles continues to diminish',
       caption='Source: Urban Institute', 
       x='Year', y='Family Income (2016 Dollars)') +
  theme_dv()
```

Disaggregating personal income from graph 3 helps display the growing wealth disparities betweens people in the tails of the income distribution. The 10th and 50th percentile families have experienced little rises in income, while the 90th percentile incomes have increased steadily since 2012. 

### Graph 5: Income

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align='center', fig.width=12, fig.height=10}
# Set working directory
setwd('/Users/ratulesrar/Desktop/data_viz_pbpl/hw2/')

median_income <- read_csv("county_incomes.csv") %>%
  select(county, state, income, state_id, county_id)

median_income %>%
  group_by(state) %>%
  ggplot(aes(x=fct_infreq(state), y=income)) +
    geom_boxplot(fill = "blue", alpha = 0.8) + 
    coord_flip() + scale_x_discrete(limits=rev(levels(income))) +
    labs(title="Median Household Income Varies Greatly from State to State",
         caption="Source: 2011-2015 American Community Survey") +
    theme(plot.title = element_text(hjust = 0.5)) + 
    labs(x="State", 
         y="Median household income (USD)",
         fill="Median household income") +
  theme_dv()
```

### Graph 6: 

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align='center', fig.width=10, fig.height=8}
# Set working directory
setwd('/Users/ratulesrar/Desktop/data_viz_pbpl/hw2/')

theme_clean_map_dv <- function(base_size = 12) {
    require(grid)
    theme_bw(base_size=11, base_family='sans') %+replace%
    theme(
        axis.title      =   element_blank(),
        axis.text       =   element_blank(),
        axis.ticks       =   element_blank(),
        panel.background    =   element_blank(),
        panel.grid      =   element_blank(),
        panel.spacing    =   unit(0,"lines"),
        plot.margin     =   unit(c(0,0,0,0),"lines"),
        legend.background=element_rect(fill='transparent', color=NA),
        legend.key=element_rect(fill='transparent', color=NA),
        plot.background=element_rect(fill='gray96', color=NA), 
        complete = TRUE
        )
    }

median_income <- read_csv("county_incomes.csv") %>%
  select(county, state, income, state_id, county_id)

states <- readOGR(dsn="cb_2016_us_state_500k", 
                  layer = "cb_2016_us_state_500k", 
                  encoding = "UTF-8", verbose = FALSE)
states_df <- fortify(states, region="GEOID")

state_incomes <- median_income %>% 
  group_by(state, state_id) %>% 
  summarize(state_median=median(income)) %>%
  mutate(id = state_id)

income_df <- full_join(states_df, state_incomes, by='id')

income_df %>%
  ggplot() +
  geom_polygon(aes(x=long, y=lat, group=group, fill=state_median),
               color='white', size=0.25) +
  guides(fill=guide_legend(title="Median Income (USD)")) +
  #scale_fill_gradient2(limits = c(0, 100000)) +
  coord_map("albers", lat0 = 29.5, lat1 = 45.5,
            xlim = c(-124.85, -66.88), ylim = c(24.4, 49.38),
            orientation = c(90, 0, -98.35)) +
  labs(title="The Northeast United States have higher median Incomes",
       caption="Source: 2011-2015 American Community Survey") +
  theme_dv() +
  theme_clean_map_dv()

```



